---
title: "Spline in a Cox Regression Model"
---

We provide an illustration of the use of restricted cubic splines in the case of breast cancer.

# The Rotterdam dataset

```{r}
#| message: false
#| warning: false

# Load libraries
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(rms)
library(emmeans)

data("gbsg", package = "survival")

# We set some variables as factors
gbsg <- gbsg %>%
  mutate(pid = factor(pid),
         # status = factor(status),
         meno = factor(meno), 
         hormon = factor(hormon), 
         grade = factor(grade))
```

For this example, we used the database constructed by the German Breast Cancer Study Group from the `survival` library.

::: panel-tabset
## Description

```{r}
#| R.options:
#|  knitr.kable.NA: ""

gbsg %>%
  mutate(status = factor(status)) %>%
  summary() %>%
  knitr::kable()
```

## Overview

```{r}
gbsg
```
:::

List of variables:

-   pid: Patient identifier
-   age: Age (years)
-   meno: Menopausal status (0 = premenopausal, 1 = postmenopausal)
-   size: Tumor size (mm)
-   grade: Tumor grade
-   nodes: Number of positive lymph nodes
-   pgr: Progesterone receptors (fmol/l)
-   er: Estrogen receptors (fmol/l)
-   hormon: Hormonal therapy (0 = no, 1= yes)
-   rfstime: Recurrence-free survival time; days to first of reccurence, death or last follow-up
-   status: 0 = alive without recurrence, 1 = recurrence or death

# Modelise age as continous variable

First let's have a look at @fig-km, showing the Kaplan-Meier curve of recurrence-free survival according to menopausal status.

```{r}
#| label: fig-km
#| fig-cap: "Kaplan-Meier curve of recurrence-free survival by menopausal status"
#| fig-align: left
#| fig-width: 8
#| fig-height: 6

ggsurvplot(survfit(Surv(rfstime, status) ~ meno, data=gbsg),
           pval = FALSE, conf.int = TRUE,
           xscale = 365.25,
           break.time.by = 365.25,
           fontsize = 4,
           xlab = "Year since ",
           risk.table = "abs_pct",  # absolute number and percentage at risk.
           risk.table.y.text = FALSE, # show bars instead of names in text annotations
           surv.median.line = "none", # Specify median survival
           ggtheme = theme_bw())
```

Now we construct a Cox proportional-hazards model with menopausal status and age.

## Model summary

```{r}
mcox <- coxph(Surv(rfstime, status) ~ meno + age, data=gbsg)

summary(mcox, scale=365.25)
```

We observe that the risk of recurrence tends to decrease with age (p-value = 0.074).

## Residuals of the model

@fig-mart-res shows the Martingale residuals of the model vs age. We can see that the linearity assumption does not look satisfied.

```{r}
#| label: fig-mart-res
#| fig-cap: "Martingale residuals and age <br> The blue area represents the 2.5 and 97.5 percentiles of the observed age distribution."
#| fig-align: left

# try also: survminer::ggcoxdiagnostics(mcox, type = "martingale")

martResb <- residuals(mcox, process = "Event")

ggplot(gbsg, aes(x=age, y=martResb)) +
  geom_hline(yintercept=0, color = "red", linetype = "dashed") + 
  geom_point(alpha = 0.2) +
  stat_smooth(method = "loess", formula = y ~ x, alpha = 0.17) + 
  annotate("rect", xmin = quantile(gbsg$age,0.025),
           xmax = quantile(gbsg$age,0.975),
           ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "#99CCCC") +
  labs(x = "Age",
       y = "Martingale residuals") + 
  theme_bw()
```

# Modelise age with splines

The restricted cubic splines are computed with the `rcs` function from the `rms` library.

We compare the following model using the likelihood-ratio test.

1.  Surv(rfstime, status) ~ meno + age : model with age as continous

2.  Surv(rfstime, status) ~ meno + rcs(age) : model with splines

```{r}
#| R.options:
#|  knitr.kable.NA: ""

mcox_spl <- coxph(Surv(rfstime, status) ~ meno + rcs(age), data=gbsg)

anova(mcox, mcox_spl, test = "LRT") %>%
  knitr::kable(digits = 3, row.names = T)
```

Comparison of the two models shows that the model with splines fits significantly better.
As suggested by the analysis of residuals (@fig-mart-res), the relationship between the risk of recurrence and age is not linear.

```{r}
#| include: false

AIC(mcox, mcox_spl)
```

```{r}
#| include: false

anova(mcox_spl)
```

Now let's have look at the summary of the model with splines.

## Model summary

```{r}
summary(mcox_spl, scale=365.25)
```

We cannot interpret the model's coefficients. 😭
We need to use graphic representations.

# Vizualisation of the models

## Compare predictions

In @fig-pred, we show the predicted recurrence-free survival probability at 3-years according to age.

```{r}
#| label: fig-pred
#| fig-cap: "Comparison of models predictions of the 3-years survival"
#| fig-align: left

# Set up
age_range <- seq(24, 90, by = 1)
t <- 3

# Cox regression
mcox_pred <- summary(survfit(mcox, newdata = data.frame(age = age_range, meno = factor(1))), 
                     times = t*365.25, scale=365.25)

# Cox regression with spline
mcox_spl_pred <- summary(survfit(mcox_spl, newdata = data.frame(age = age_range, meno = factor(1))), 
                     times = t*365.25, scale=365.25)

# Merge
mcox_pred_all <- bind_cols(age = rep(age_range, 2),
                           Model = rep(c("Cox", "Cox_spline"), each = length(age_range)),
                           Time = t,
                           Surv_Prob = c(mcox_pred$surv, mcox_spl_pred$surv), 
                           Lower = c(mcox_pred$lower, mcox_spl_pred$lower), 
                           Upper = c(mcox_pred$upper, mcox_spl_pred$upper))


ggplot(mcox_pred_all, aes(x = age, y = Surv_Prob)) +
  geom_line(aes(color = Model), linewidth = 1) +
  geom_ribbon(aes(ymin=Lower, ymax=Upper, fill = Model), alpha=.2) +
  scale_discrete_manual(c("color", "fill"),
                        values=c("Cox" = "cyan4", "Cox_spline" = "goldenrod2")) +
  labs(x = "Age", y = "Survival at 3 years", color = "", fill = "") +
  theme_bw() +
  theme(legend.title = element_text(size = 15, face = 'bold'),
        legend.text = element_text(size = 15),
        text = element_text(size=15), legend.position="bottom")
```

```{r}
#| include: false
#| label: fig-pred-ci
#| fig-cap: "Comparison of model predictions"
#| fig-subcap: 
#|   - "Linear regression"
#|   - "Linear regression with spline"
#| layout-ncol: 2
#| fig-align: left

ggeffects::ggpredict(mcox, terms = "age", type = "survival") %>%
  plot()

ggeffects::ggpredict(mcox_spl, terms = "age", type = "survival") %>%
  plot()
```

## Contrast analysis

In @fig-ctrs, we compare the Hazard Ratio of recurrence with a reference age (e.g. 45).

```{r}
#| label: fig-ctrs
#| fig-cap: "Contrast analysis in both models"
#| fig-subcap: 
#|   - "Cox regression"
#|   - "Cox regression with spline"
#| layout-ncol: 2
#| fig-align: left

# Set up
ref <- 45

# Cox regression
mcox_emm <- emmeans(mcox, ~ age, at = list(age = age_range), type = "response")

mcox_ctrs <- emmeans::contrast(mcox_emm, "trt.vs.ctrl", ref = paste0("age", ref), 
                               infer = T,  adjust = "none")
  
mcox_ctrs %>%
  as.data.frame() %>%
  bind_cols("age" = age_range[age_range != ref]) %>% 
  ggplot(., aes(x=age, y=ratio)) +
  geom_hline(yintercept = 1, color = "red") +
  geom_line() +
  geom_ribbon(aes(ymin=asymp.LCL, ymax=asymp.UCL), alpha = .2, linetype = 0) +
  labs(x = "age", y = "Hazard Ratio") +
  theme_bw()

# Cox regression with spline
mcox_spl_emm <- emmeans(mcox_spl, ~ age, at = list(age = age_range), type = "response")

mcox_spl_ctrs <- emmeans::contrast(mcox_spl_emm, "trt.vs.ctrl", ref = paste0("age", ref), 
                                   infer = T,  adjust = "none")
  
mcox_spl_ctrs %>%
  as.data.frame() %>%
  bind_cols("age" = age_range[age_range != ref]) %>% 
  ggplot(., aes(x=age, y=ratio)) +
  geom_hline(yintercept = 1, color = "red") +
  geom_line() +
  geom_ribbon(aes(ymin=asymp.LCL, ymax=asymp.UCL), alpha = .2, linetype = 0) +
  labs(x = "age", y = "Hazard Ratio") +
  theme_bw()
```

```{r}
#| R.options:
#|  knitr.kable.NA: ""

mcox_spl_ctrs[mcox_spl_ctrs@grid == "age30 - age45"] %>%
  knitr::kable(digits = 3)
```


