---
title: "Spline in a Linear Regression Model"
---

We provide an illustration of the use of restricted cubic splines in the case of diabetes.

# The Pima Indians Diabetes Database

```{r}
#| message: false
#| warning: false

# Load libraries
library(dplyr)
library(ggplot2)
library(broom)
library(mlbench)
library(rms)
library(emmeans)
library(ggeffects)

data("PimaIndiansDiabetes2", package = "mlbench")

# We remove women with missing glucose concentration
PimaIndiansDiabetes2 <- PimaIndiansDiabetes2 %>%
  filter(!is.na(glucose))
```

For this example, we used the Pima Indians Diabetes Database from the `mlbench` library.

The dataset consists of 9 demographic and clinical variables collected in women 21 years and older of [Pima Indian](https://en.wikipedia.org/wiki/Akimel_O%27odham) heritage. This population is known to have a high prevalence of type 2 diabetes.

::: panel-tabset
## Description

```{r}
#| R.options:
#|  knitr.kable.NA: ""

summary(PimaIndiansDiabetes2) %>%
  knitr::kable()
```

## Overview

```{r}
PimaIndiansDiabetes2
```
:::

List of variables:

-   pregnant : Number of times pregnant
-   glucose : Plasma glucose concentration (glucose tolerance test)
-   pressure : Diastolic blood pressure (mm Hg)
-   triceps : Triceps skin fold thickness (mm)
-   insulin : 2-Hour serum insulin (mu U/ml)
-   mass : Body mass index (weight in kg/(height in m)^2^)
-   pedigree : Diabetes pedigree function
-   age : Age (years)
-   diabetes : Test for diabetes (neg/pos)

# Modelise age as continous variable

First let's have a look at @fig-age-glc, showing the relationship between plasma glucose concentration and age.

```{r}
#| label: fig-age-glc
#| fig-cap: "Plasma glucose concentration according to age <br> The blue curve is a local smoothing (loess) fitting the observed data."
#| fig-align: left

ggplot(PimaIndiansDiabetes2, aes(age, glucose)) +
  geom_point(alpha = .2) +
  stat_smooth(method = "loess", formula = y ~ x, alpha = 0.17) +
  theme_bw()
```

Now we construct a linear regression model of the relationship between glucose and age.

## Model summary

```{r}
lin_reg <- lm(glucose ~ age, data = PimaIndiansDiabetes2)

summary(lin_reg)
```

We observe that glucose concentration increases significantly with age (β=0.69 ± 0.09, p-value \< 0.001).

## Residuals of the model

@fig-lm-res shows the residuals of the model vs age. We can see that the linearity assumption looks satisfied.

```{r}
#| label: fig-lm-res
#| fig-cap: "Residuals and age <br> The blue area represents the 2.5 and 97.5 percentiles of the observed age distribution."
#| fig-align: left

lin_reg.pred <- augment(lin_reg)
  
ggplot(lin_reg.pred, aes(x=age, y=.resid)) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = .2) +
  stat_smooth(method = "loess", formula = y ~ x, alpha = 0.2) + 
  labs(x = "Age", y = "Raw residuals") + 
  annotate("rect", 
           xmin = quantile(lin_reg.pred$age, .025),
           xmax = quantile(lin_reg.pred$age, .975),
           ymin = -Inf, ymax = Inf, alpha = .2, fill = "#99CCCC") +
  theme_bw()
```

# Modelise age with splines

The restricted cubic splines are computed with the `rcs` function from the `rms` library.

We compare the following model using the likelihood-ratio test.

1.  glucose ~ age : model with age as continous

2.  glucose ~ rcs(age) : model with splines

```{r}
#| R.options:
#|  knitr.kable.NA: ""

lin_reg_spl <- lm(glucose ~ rcs(age), data = PimaIndiansDiabetes2)

anova(lin_reg, lin_reg_spl, test = "LRT") %>%
  knitr::kable(digits = 3, row.names = T)
```

Comparison of the two models shows that the relationship between glucose and age is linear as suggested by the analysis of residuals (@fig-lm-res).

```{r}
#| include: false

AIC(lin_reg, lin_reg_spl)
```

```{r}
#| include: false

anova(lin_reg_spl)
```

Since the model with continuous age is valid, it is not necessary to use splines.

However, for illustrative purposes, let's look at the summary of the model with splines.

## Model summary

```{r}
summary(lin_reg_spl)
```

We cannot interpret the model's coefficients. 😭
We need to use graphic representations.

# Vizualisation of the models

## Compare predictions

@fig-pred show the predictions of the models.

```{r}
#| label: fig-pred
#| fig-cap: "Comparison of models predictions"
#| fig-align: left

lin_reg_spl.pred <- predict(lin_reg_spl, type = "response", interval = "confidence")

ggplot(lin_reg.pred, aes(x = age)) +
  geom_point(aes(y = glucose), alpha = .2) +
  stat_smooth(aes(y = glucose, color = "LOESS"), method = "loess", formula = y ~ x, se = FALSE,) + 
  geom_line(aes(y = .fitted, color = "Linear regression"), linewidth = 1) +
  geom_line(aes(y = lin_reg_spl.pred[,c("fit")], color = "Linear regression with spline"), linewidth = 1) +
  scale_color_manual(values=c("LOESS" = "#CC3333", 
                              "Linear regression" = "cyan4", 
                              "Linear regression with spline" = "goldenrod2")) +
  labs(x = "age", y = "glucose", color = "") +
  theme_bw() + 
  theme(legend.position="bottom")
```

@fig-pred-ci show the predictions with confidence band of the models.

```{r}
#| label: fig-pred-ci
#| fig-cap: "Comparison of models predictions"
#| fig-subcap: 
#|   - "Linear regression"
#|   - "Linear regression with spline"
#| layout-ncol: 2
#| fig-align: left

ggpredict(lin_reg, terms = "age") %>%
  plot(add.data = T)

ggpredict(lin_reg_spl, terms = "age") %>%
  plot(add.data = T)
```

## Contrast analysis

In @fig-ctrs, we compare the difference in plasma glucose concentration with a reference age (e.g. 30).

```{r}
#| label: fig-ctrs
#| fig-cap: "Contrast analysis in both models"
#| fig-subcap: 
#|   - "Linear regression"
#|   - "Linear regression with spline"
#| layout-ncol: 2
#| fig-align: left

# Set up
age_range <- with(PimaIndiansDiabetes2, seq(min(age), max(age), by = .5))
ref <- 30

# Linear regression
lin_reg.emm <- emmeans(lin_reg, ~ age, at = list(age = age_range), type = "response")

lin_reg.ctrs <- emmeans::contrast(lin_reg.emm, "trt.vs.ctrl", ref = paste0("age", ref), 
                                  infer = T,  adjust = "none")

lin_reg.ctrs %>%
  as.data.frame() %>%
  bind_cols("age" = age_range[age_range != ref]) %>% 
  ggplot(., aes(x=age, y=estimate)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_line() +
  geom_ribbon(aes(ymin=lower.CL, ymax=upper.CL), alpha = .2, linetype = 0) +
  labs(x = "age", y = "Glucose (Difference)") +
  theme_bw()

# Linear regression with spline
lin_reg_spl.emm <- emmeans(lin_reg_spl, ~ age, at = list(age = age_range), type = "response")

lin_reg_spl.ctrs <- emmeans::contrast(lin_reg_spl.emm, "trt.vs.ctrl", ref = paste0("age", ref), 
                                      infer = T,  adjust = "none")
  
lin_reg_spl.ctrs %>%
  as.data.frame() %>%
  bind_cols("age" = age_range[age_range != ref]) %>% 
  ggplot(., aes(x=age, y=estimate)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_line() +
  geom_ribbon(aes(ymin=lower.CL, ymax=upper.CL), alpha = .2, linetype = 0) +
  labs(x = "age", y = "Glucose (Difference)") +
  theme_bw()
```
