---
title: "Handling Continuous Variables"
bibliography: references.bib
---

When building a regression model, several strategies exist for continuous variables:

-   Categorizing the variable into two or more modalities
-   Use the variable as it is
-   Modeling the functional form

# Dichotomization

See [@Royston2006].

# Functional form

## Variable Transformation

Polynomials

Fractional polynomials

## Spline

```{r}
library(splines)
library(dplyr)
library(ggplot2)
library(patchwork)

set.seed(1)
```

```{r}
data <- tibble(x = seq(from = 0, to = 6, by = .025)) %>%
  mutate(y = sin(2*x) + x -.1*x^2 + 2 + rnorm(length(x), sd = .3))

g_template <- ggplot(data, aes(x = x, y = y)) +
  geom_point() +
  geom_vline(xintercept = c(0, 6), linetype = "dashed", color = "red") +
  theme_bw()


g_template + 
  geom_smooth(method = lm, formula = y ~ bs(x, degree = 1, knots = c(1, 2.5, 4, 5.7))) +
  geom_vline(xintercept = c(1, 2.5, 4, 5.7), linetype = "dashed") +
  ggtitle("Linear spline (4 knots)") +
  g_template + geom_smooth(method = lm, formula = y ~ bs(x, degree = 2, knots = c(1, 2.5, 4, 5.7))) +
  geom_vline(xintercept = c(1, 2.5, 4, 5.7), linetype = "dashed") +
  ggtitle("Quadratic spline (4 knots)") +
  g_template + geom_smooth(method = lm, formula = y ~ bs(x, degree = 3, knots = c(1, 2.5, 4, 5.7))) +
  geom_vline(xintercept = c(1, 2.5, 4, 5.7), linetype = "dashed") +
  ggtitle("Cubic spline (4 knots)")

g_template + 
  geom_smooth(method = lm, formula = y ~ bs(x, degree = 3, knots = c(3))) +
  geom_vline(xintercept = c(3), linetype = "dashed") +
  ggtitle("Cubic spline (1 knots)") +
  g_template + geom_smooth(method = lm, formula = y ~ bs(x, degree = 3, knots = c(1, 2.5, 4, 5.7))) +
  geom_vline(xintercept = c(1, 2.5, 4, 5.7), linetype = "dashed") +
  ggtitle("Cubic spline (4 knots)") +
  g_template + geom_smooth(method = lm, formula = y ~ bs(x, degree = 3, knots = seq(from = 0.1, to = 5.9, by = .2))) +
  geom_vline(xintercept = seq(from = 0.1, to = 5.9, by = .2), linetype = "dashed") +
  ggtitle("Cubic spline (30 knots)")
```

Example of splines implementations in regression models :

-   Linear regression
-   Logistic regression
-   Cox proportional-hazards model
-   Linear mixed model
